Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-04-26 MIMOSA.tex
Some changes to the model section

\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}
\setromanfont[Mapping=tex-text]{Times New Roman}
\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Gill Sans}
\setmonofont[Scale=MatchLowercase]{Andale Mono}

\title{Combinatorial Mixture Models for Single Cell Assays}
\author[1]{Greg Finak}
\author[2]{SC De Rosa}
\author[1]{Raphael Gottardo}

@@

\section{Introduction}
Single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple variables simultaneously on individual cells in a heterogeneous mixture such as whole blood. These variables are used to classify individual cells in the mixture into more homogeneous sub--populations based on phenotypic or functional differences. Such single--cell assays are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify individuals whose immune system responds to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T--cels via antigen presenting cells. While not all T--cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future\cite{McKinstry:2010ei}. These antigen--specific T--cell subpopulations a very small fraction of the total number of CD4 and CD8 T--cells. The ICS assay measures the number of antigen--specific T--cells in whole blood by measuring cytokine production in response to activation following stimulation by an antigen that was present in the original vaccine. Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8) and functional markers (cytokines) of the cell subpopulations of interest~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences. Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw, Bendall:2011wf}.

The ICS assay in flow cytometry is just one example of the applications of single--cell technologies to immunology. Single--cell gene expression technologies such as Fluidigm are enabling researchers to measure the expression of 96 genes in 96 single cells from a homogeneous population sorted by flow cytometry\cite{Jang:2011uj}. This technology has been used to identify signatures of immune cell sub--populations that correlate with different types of vaccine~\cite{Flatz:2011jb}. To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements. However, the proportion of single cells expressing individual genes also carries information and should be evaluated.

Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test)~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}. These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}.

@@

The framework developed in this paper addresses these issues explicitly. We present a combinatorial mixture model framework for the analysis of single--cell assay data where multivariate measurement are made on individual cells. The model is used to identify observations where a significant difference exists between paired treatment and control samples with respect to the number of cells expressing  different combinations of proteins, genes, or other measured properties of the cells. Importantly, our approach shares information across observations by means of prior distributions on the responding, stimulated samples and the non--responding samples thereby increasing sensitivity and specificity to detect responding cell populations. The cell counts from the stimulated and unstimulated experiments are modelled independently, and different combinations of markers are represented as different mixture components in the framework. This approach allows us to omit certain combinations of markers that are not observable (i.e., two cytokines may never be co--expressed), or to explicitly represent combinations that are of interest (i.e., we can explicitly represent and test for a difference in a pair of cytokines in an experiment measuring many cytokines). This is a flexible approach that avoids the some of the drawbacks of global tests when only a few of the many measured markers show differences~\cite{Nason:2006dx}.

\section{Data structure and notation} 
We analyzed two data sets from two different immunological single--cell assays typically used in vaccine trials, one flow cytometry data set, and one single--cell gene expression data set.

\textit{Flow cytometry}: This ICS data is from a phase--I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines in two cell populations from 20 individuals at two time points (zero and 28 days post--vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}. The statistical analysis of responder and non--responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}. The goal of this data set was to assess response rates of CD4 and CD8 T--cell populations to different antigens.

@@

In the remainder of the paper, we use the following notation to describe our model. We observe cell counts from $P$ subjects in two conditions (stimulated and unstimulated). Each cell can either be positive or negative for a marker. We let $i$ index the subjects and $k$ index the cytokines. The subscripts $s$,$u$ refer to the cell counts from stimulated and unstimulated conditions, respectively. The total number of cells is denoted by $N$, and given in our model, while $n$, the number of positive cells is part of the observe data. Thus for a given cytokine, $N_{si}$ denotes the total number of cells collected for subject $i$ in the stimulated condition and $n_{si}$ denotes how many of those cells were cytokine positive. Similarly, $N_{ui}$, $n_{ui}$ represent the total, and cytokine positive cell counts in the unstimulated condition for the same subject. For each cytokine and each stimulation, given the total cell counts, the observed data is a matrix of size $2 \times P$, where the $i$'th row is a subject, $\mathbf{y}_i = <n_{si},n_{ui}>$.

@@

  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Cytokine}\\
 & Negative & Positive \\ 
  \hline
Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\ 
Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\ 
   \hline
\end{tabular}
}
\end{table}
We consider a marker to be differentially expressed between conditions if the number of cells in the stimulated samples is significantly different from the number of cells in the control sample. The model is described below.

\subsection{Beta--Binomial Model for One Marker}
The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as  binomial random variables.
\begin{align}
(n_{si}|p_{ui}) \sim \mathrm{Bin}(N_{si},p_{ui});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null};&\text{if $i$ is a non--responder}\\
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui});&\text{if $i$ is a responder}\label{eq:alternate}
 \end{align}

Where $p_{si}$, $p_{ui}$ are the unobserved proportions. Model~\eqref{eq:null} represents the \textit{null} hypothesis where there is no difference between the stimulation and the control, and thus no difference between the proportion of cytokine positive cells in stimulated and control samples. Model~\eqref{eq:alternate} represents the  alternative hypothesis where there is a difference in cytokine response between the stimulation and the control, and thus a difference in the proportion of cells between stimulated and control samples. In the case of no response, the counts for the stimulated and unstimulated samples are independent draws from the same  binomial distribution.

\subsection{Prior}
Our model shares information across responders and non--responders, respectively. This is done by placing Beta priors on the $p_{si}$, $p_{ui}$, across subjects:

 \begin{align}
 p_{si} &\sim \mathrm{Beta}(\alpha_s,\beta_s);&
 p_{ui} &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:prior}
 \end{align}

Additional sharing of information is done through the $\alpha_u, \beta_u$ hyper--parameters, which are common for the null and alternative models. 
%If $p_{si}=p_{ui}$ we assume that $\alpha_s=\alpha_u$ and $\beta_s=\beta_u$, thus sharing the hyper--parameters between the null and alternative model for the unstimulated samples, such that $\beta_u,\alpha_u$ hyper--parameters are equal for both the stimulated and unstimulated models. Given this formulation, the posterior probability of the data given that it is generated by model \eqref{eq:null}, is:
\subsection{Posetrior}
\subsubsection{Null Model}
Given the above formulation, the posterior for the null model is Beta--binomial, with marginal likelihood (integrating out $p_u$):

>>>>

%\usepackage{fontspec,xltxtra,xunicode}
%\defaultfontfeatures{Mapping=tex-text}
%\setromanfont[Mapping=tex-text]{Times New Roman}
%\setsansfont[Scale=MatchLowercase,Mapping=tex-text]{Gill Sans}
%\setmonofont[Scale=MatchLowercase]{Andale Mono}

\title{Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies}
\author[1]{Greg Finak}
\author[2]{SC De Rosa}
\author[1]{Raphael Gottardo}

@@

\section{Introduction}
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers). Recently, DVS Sciences Inc. introduced a new cytometry technology named cytometry Time-Of-Flight (CyTOF) that uses rare isotopes instead of fluorochromes (as in standard FCM), as antibody tags to label single cells. CyTOF drastically extends the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell. Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize cell poly-functionality. A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes6. While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability. This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data. Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general. For this reason, single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple variables simultaneously on individual cells in a heterogeneous mixture such as whole blood. These variables are used to classify individual cells in the mixture into more homogeneous sub--populations based on phenotypic or functional differences. Such single--cell assays are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

A motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify individuals' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T--cels via antigen presenting cells. While not all T--cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce a variety of cytokines, further promoting the immune response. After activation, this antigen--specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen again in the future \cite{McKinstry:2010ei}. These antigen--specific T--cell subpopulations constitute a very small fraction of the total number of CD4 and CD8 T--cells. The ICS assay measures the number of antigen--specific T--cells in whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8) and functional markers (cytokines) of the cell subpopulations of interest~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences. Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw, Bendall:2011wf}.

%The ICS assay in flow cytometry is just one example of the applications of single--cell technologies to immunology. Single--cell gene expression technologies such as Fluidigm are enabling researchers to measure the expression of 96 genes in 96 single cells from a homogeneous population sorted by flow cytometry\cite{Jang:2011uj}. This technology has been used to identify signatures of immune cell sub--populations that correlate with different types of vaccine~\cite{Flatz:2011jb}. To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements. However, the proportion of single cells expressing individual genes also carries information and should be evaluated.

Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test)~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}. These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}. In the context of single-cell gene expression, very few methods have been proposed. To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements. However, the proportion of single cells expressing individual genes also carries information and should be evaluated.

@@

The framework developed in this paper addresses these issues explicitly. We present a multinomial-Dirichlet combinatorial mixture model framework for the analysis of single--cell assay data where multivariate measurement are made on individual cells. The model is used to identify observations where a significant difference exists between paired treatment and control samples with respect to the number of cells expressing  different combinations of proteins, genes, or other measured properties of the cells. Importantly, our approach shares information across individuals (or subjects) by means of a Dirichlet prior distribution placed on the unknown cell population proportions of the mutinomial likelihood, which help increase sensitivity and specificity to detect rare antigen specific responding cell populations. The cell counts from the stimulated and unstimulated experiments are modeled independently, and different combinations of markers are represented as different mixture components in the framework. This approach allows us to omit certain combinations of markers that are not observable (i.e., two cytokines may never be co--expressed), or to explicitly represent combinations that are of interest (i.e., we can explicitly represent and test for a difference in a pair of cytokines in an experiment measuring many cytokines). This is a flexible approach that avoids the some of the drawbacks of global tests when only a few of the many measured markers show differences~\cite{Nason:2006dx}.

\section{Data structure and notation} 
In this paper we consider two different immunological single-cell assays typically used in vaccine trials, one flow cytometry data set, and one single-cell gene expression data set.

\textit{Flow cytometry}: This ICS data is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}. The statistical analysis of responder and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}. The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.

@@

In the remainder of the paper, we use the following notation to describe our model. We observe cell counts from $I$ individuals in two conditions (stimulated and unstimulated). Each cell can either be positive or negative for a marker. We let $i$ index the individual and $k$ index the measured markers, \textit{i.e.} cytokine for the ICS data and genes for the ICS data. The subscripts $s$,$u$ refer to the cell counts from stimulated and unstimulated conditions, respectively. For a given individual, $p$, the total number of cells is denoted by $N_p$, while $n_p$, the number of positive cells is part of the observe data. Thus for a given cytokine, $N_{si}$ denotes the total number of cells collected for subject $i$ in the stimulated condition and $n_{si}$ denotes how many of those cells were cytokine positive. Similarly, $N_{ui}$, $n_{ui}$ represent the total, and cytokine positive cell counts in the unstimulated condition for the same subject. For each cytokine and each stimulation, given the total cell counts, the observed data is a matrix of size $2 \times P$, where the $i$'th row is a subject, $\mathbf{y}_i = <n_{si},n_{ui}>$.

@@

  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Cytokine}\\
 & Negative & Positive \\ 
  \hline
Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\ 
Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\ 
   \hline
\end{tabular}
}
\end{table}
We consider a marker to be differentially expressed between conditions if the proportion of cells in the stimulated samples is different from the number of cells in the control sample. The model is described below.

\subsection{Beta--Binomial Model for One Marker}
The positive cell counts for stimulated and unstimulated samples from the same individual are modelled as binomial random variables, as follows,
\begin{align}
(n_{si}|p_{ui}) \sim \mathrm{Bin}(N_{si},p_{ui});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null};&\text{if $i$ is a non--responder}\\
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui});&\text{if $i$ is a responder}\label{eq:alternate}
 \end{align}

where $p_{si}$, $p_{ui}$ are the unobserved proportions for the stimulated and un-stimulated samples. Under the null hypothesis, \textit{i.e.} Equation~\ref{eq:null}, there is no difference between the stimulation and control samples, and the proportions are equal. Model~\eqref{eq:alternate} represents the  alternative hypothesis where there is a difference in proportions between the stimulation and control samples.

\subsection{Priors}
Our model shares information across all individuals using an exchangeable Beta priors on the unknown proportions, as follows, 
 \begin{align}
(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:prior-null}\\
(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u) \label{eq:prior-alternative}
 \end{align}
where $z_i$ is an indicator variable equal to one if individual $i$ is a responder, i.e.  $p_{si}\ne p_{ui}$ (or $p_{si}> p_{ui}$), and 0 otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals. Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models. Under the null model, the stimulated condition should behave just like the unstimulated one, and thus
the stimulated response rates (\textit{i.e.} $p_{si}$'s) should follow the same prior distribution. Finally, we assume that the $z_i$'s are independent and identically distributed as Bernoulli with probability $w$.
Treating the $z_i$'s as missing data, the unknown parameter vector $\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can easily be estimated via an Empirical-Bayes fashion using and Expectation-Maximization algorithm as described in Section \ref{sec:estimation}.

\subsection{Parameter estimation}
The log complete data likelihood can be written as 
\begin{equation}
l(\theta|y)=\sum_i z_i l_0(\alpha_u, \beta_u|y_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|y_i)
\end{equation}
where $l_0$ and $l_1$ are the marginal likelihoods for the null and alternative models obtained after integrating out the prior for $\alpha_u, \beta_u, \alpha_s$, and $\beta_s$.

\subsubsection{Null Model}
Given the above formulation, the posterior for the null model is Beta--binomial, with marginal likelihood (integrating out $p_u$):
